---
title: "Assignment 8.2"
author: "Scott Breitbach"
date: "7/19/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/micha/OneDrive/Documents/GitHub/DSC520/")
# read files
library(readr)
# plot data
library(ggplot2)
# knn() function
library(class)
# CrossTable() function
library(gmodels)
```

## Assignment 8.2: Introduction to Machine Learning

```{r read files, echo = TRUE, message = FALSE}
BCdata <- read_csv("completed/Week8/data/binary-classifier-data.csv")
BCdata$label <- factor(BCdata$label)
summary(BCdata)
str(BCdata)

TCdata <- read_csv("completed/Week8/data/trinary-classifier-data.csv")
TCdata$label <- factor(TCdata$label)
summary(TCdata)
str(TCdata)
```

### a. Plot the data from each dataset using a scatter plot.

```{r plot binary, echo = TRUE}
ggplot(data = BCdata, mapping = aes(x, y, color = label)) + geom_point() + ggtitle("Binary Data Plot")
```

```{r plot trinary, echo = TRUE}
ggplot(data = TCdata, mapping = aes(x, y, color = label)) + geom_point() + ggtitle("Trinary Data Plot")
```

### b. Determine which points are nearest by calculating the Euclidean distance between two points.

#### Fit a k nearest neighbors model for each dataset for k=3, k=5, k=10, k=15, k=20, and k=25. Compute the accuracy of the resulting models for each value of k. 

```{r setup test train, echo = TRUE}
## Set up testing and training groups from data sets
set.seed(42)

indBC <- sample(2, nrow(BCdata), replace=TRUE, prob=c(0.67, 0.33))
BC.train <- BCdata[indBC==1, 2:3]
BC.test <- BCdata[indBC==2, 2:3]
BC.trainLabels <- BCdata[indBC==1,1]
BC.trainLabels <- BC.trainLabels$label
BC.testLabels <- BCdata[indBC==2,1]
BC.testLabels <- BC.testLabels$label

indTC <- sample(2, nrow(TCdata), replace=TRUE, prob=c(0.67, 0.33))
TC.train <- TCdata[indTC==1, 2:3]
TC.test <- TCdata[indTC==2, 2:3]
TC.trainLabels <- TCdata[indTC==1,1]
TC.trainLabels <- TC.trainLabels$label
TC.testLabels <- TCdata[indTC==2,1]
TC.testLabels <- TC.testLabels$label
```

```{r knn k3, echo = TRUE}
BCpred.3 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=3)
AccBC.3 <- sum(BC.testLabels==BCpred.3)/length(BCpred.3)
"Accuracy: Binary, k=3" 
round(AccBC.3 * 100, digits = 1)

TCpred.3 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=3)
AccTC.3 <- sum(TC.testLabels==TCpred.3)/length(TCpred.3)
"Accuracy: Trinary, k=3"
round(AccTC.3 * 100, digits = 1)
```

```{r knn k5, echo = TRUE}
BCpred.5 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=5)
AccBC.5 <- sum(BC.testLabels==BCpred.5)/length(BCpred.5)
"Accuracy: Binary, k=5" 
round(AccBC.5 * 100, digits = 1)

TCpred.5 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=5)
AccTC.5 <- sum(TC.testLabels==TCpred.5)/length(TCpred.5)
"Accuracy: Trinary, k=5"
round(AccTC.5 * 100, digits = 1)
```

```{r knn k10, echo = TRUE}
BCpred.10 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=10)
AccBC.10 <- sum(BC.testLabels==BCpred.10)/length(BCpred.10)
"Accuracy: Binary, k=10" 
round(AccBC.10 * 100, digits = 1)

TCpred.10 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=10)
AccTC.10 <- sum(TC.testLabels==TCpred.10)/length(TCpred.10)
"Accuracy: Trinary, k=10"
round(AccTC.10 * 100, digits = 1)
```

```{r knn k15, echo = TRUE}
BCpred.15 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=15)
AccBC.15 <- sum(BC.testLabels==BCpred.15)/length(BCpred.15)
"Accuracy: Binary, k=15" 
round(AccBC.15 * 100, digits = 1)

TCpred.15 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=15)
AccTC.15 <- sum(TC.testLabels==TCpred.15)/length(TCpred.15)
"Accuracy: Trinary, k=15"
round(AccTC.15 * 100, digits = 1)
```

```{r knn k20, echo = TRUE}
BCpred.20 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=20)
AccBC.20 <- sum(BC.testLabels==BCpred.20)/length(BCpred.20)
"Accuracy: Binary, k=20" 
round(AccBC.20 * 100, digits = 1)

TCpred.20 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=20)
AccTC.20 <- sum(TC.testLabels==TCpred.20)/length(TCpred.20)
"Accuracy: Trinary, k=20"
round(AccTC.20 * 100, digits = 1)
```

```{r knn k25, echo = TRUE}
BCpred.25 <- knn(train = BC.train, test = BC.test, cl = BC.trainLabels, k=25)
AccBC.25 <- sum(BC.testLabels==BCpred.25)/length(BCpred.25)
"Accuracy: Binary, k=25" 
round(AccBC.25 * 100, digits = 1)

TCpred.25 <- knn(train = TC.train, test = TC.test, cl = TC.trainLabels, k=25)
AccTC.25 <- sum(TC.testLabels==TCpred.25)/length(TCpred.25)
"Accuracy: Trinary, k=25"
round(AccTC.25 * 100, digits = 1)
```

#### Plot the results in a graph where the x-axis is the different values of k and the y-axis is the accuracy of the model.

```{r plot k vs acc, echo = TRUE}
## Build the data frame
kValues <- c(3, 5, 10, 15, 20, 25)

# Binary data
pctAcc <- c(AccBC.3, AccBC.5, AccBC.10, AccBC.15, AccBC.20, AccBC.25) * 100
label <- c("Binary")
BCAccSet <- data.frame(label, kValues, pctAcc)

# Trinary data
pctAcc <- c(AccTC.3, AccTC.5, AccTC.10, AccTC.15, AccTC.20, AccTC.25) * 100
label <- c("Trinary")
TCAccSet <- data.frame(label, kValues, pctAcc)

# Combine data sets
dataSet <- rbind(BCAccSet, TCAccSet)

# Plot the data
ggplot(data = dataSet, mapping = aes(kValues, pctAcc, color = label)) + 
  geom_point() + labs(x = "k Values", y = "% Accuracy", 
                      title = "Percent Accuracy of Different k Values 
                      for Binary and Trinary Data Sets")
```

### c. Looking back at the plots of the data, do you think a linear classifier would work well on these datasets?

Looking at the plots of the Binary and Trinary data sets, it seems pretty obvious that linear classifiers would not be useful in determining to which groups the individual data points would belong. They are clustered together, which is helpful in the nearest neighbors algorithm, but the clusters are too interspersed with each other to be divided by any sort of reasonable line.

