---
title: "Assignment 8.3"
author: "Scott Breitbach"
date: "7/20/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/micha/OneDrive/Documents/GitHub/DSC520/")
# read files
library(readr)
# plot data
library(ggplot2)
set.seed(42)
```

## Assignment 8.3: Clustering

```{r read file, echo = TRUE, message = FALSE}
clusterDF <- read_csv("completed/Week8/data/clustering-data.csv")
head(clusterDF)
summary(clusterDF)
```

### a. Plot the dataset using a scatter plot.

```{r plot data, echo = TRUE}
ggplot(data = clusterDF, mapping = aes(x, y)) + geom_point()
```

### b. Fit the dataset using the k-means algorithm from k=2 to k=12. Create a scatter plot of the resultant clusters for each value of k.

```{r kmeans k2, echo = TRUE}
clusters.2 <- kmeans(clusterDF, 2)
clusterDF$k2 <- as.factor(clusters.2$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k2)) + geom_point()
```

```{r kmeans k3, echo = TRUE}
clusters.3 <- kmeans(clusterDF, 3)
clusterDF$k3 <- as.factor(clusters.3$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k3)) + geom_point()
```

```{r kmeans k4, echo = TRUE}
clusters.4 <- kmeans(clusterDF, 4)
clusterDF$k4 <- as.factor(clusters.4$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k4)) + geom_point()
```

```{r kmeans k5, echo = TRUE}
clusters.5 <- kmeans(clusterDF, 5)
clusterDF$k5 <- as.factor(clusters.5$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k5)) + geom_point()
```

```{r kmeans k6, echo = TRUE}
clusters.6 <- kmeans(clusterDF, 6)
clusterDF$k6 <- as.factor(clusters.6$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k6)) + geom_point()
```

```{r kmeans k7, echo = TRUE}
clusters.7 <- kmeans(clusterDF, 7)
clusterDF$k7 <- as.factor(clusters.7$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k7)) + geom_point()
```

```{r kmeans k8, echo = TRUE}
clusters.8 <- kmeans(clusterDF, 8)
clusterDF$k8 <- as.factor(clusters.8$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k8)) + geom_point()
```

```{r kmeans k9, echo = TRUE}
clusters.9 <- kmeans(clusterDF, 9)
clusterDF$k9 <- as.factor(clusters.9$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k9)) + geom_point()
```

```{r kmeans k10, echo = TRUE}
clusters.10 <- kmeans(clusterDF, 10)
clusterDF$k10 <- as.factor(clusters.10$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k10)) + geom_point()
```

```{r kmeans k11, echo = TRUE}
clusters.11 <- kmeans(clusterDF, 11)
clusterDF$k11 <- as.factor(clusters.11$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k11)) + geom_point()
```

```{r kmeans k12, echo = TRUE}
clusters.12 <- kmeans(clusterDF, 12)
clusterDF$k12 <- as.factor(clusters.12$cluster)

ggplot(clusterDF, mapping = aes(x, y, color = k12)) + geom_point()
```

### c. To calculate the average distance from the center of each cluster, compute the distance of each data point to the center of the cluster it is assigned to and take the average value of all of those distances.

```{r dist function, echo = TRUE}
myDist<-function(p1,p2) sqrt((p1[,1]-p2[,1])^2+(p1[,2]-p2[,2])^2)
```

```{r k2 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.2$cluster==1,],clusters.2$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.2$cluster==2,],clusters.2$centers[2,,drop=FALSE])
# Get average distance
k2.avdist <- sum(dist.1, dist.2)/nrow(clusterDF)
k2.avdist
```

```{r k3 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.3$cluster==1,],clusters.3$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.3$cluster==2,],clusters.3$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.3$cluster==3,],clusters.3$centers[3,,drop=FALSE])
# Get average distance
k3.avdist <- sum(dist.1, dist.2, dist.3)/nrow(clusterDF)
k3.avdist
```

```{r k4 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.4$cluster==1,],clusters.4$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.4$cluster==2,],clusters.4$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.4$cluster==3,],clusters.4$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.4$cluster==4,],clusters.4$centers[4,,drop=FALSE])
# Get average distance
k4.avdist <- sum(dist.1, dist.2, dist.3, dist.4)/nrow(clusterDF)
k4.avdist
```

```{r k5 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.5$cluster==1,],clusters.5$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.5$cluster==2,],clusters.5$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.5$cluster==3,],clusters.5$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.5$cluster==4,],clusters.5$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.5$cluster==5,],clusters.5$centers[5,,drop=FALSE])
# Get average distance
k5.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5)/nrow(clusterDF)
k5.avdist
```

```{r k6 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.6$cluster==1,],clusters.6$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.6$cluster==2,],clusters.6$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.6$cluster==3,],clusters.6$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.6$cluster==4,],clusters.6$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.6$cluster==5,],clusters.6$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.6$cluster==6,],clusters.6$centers[6,,drop=FALSE])
# Get average distance
k6.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6)/nrow(clusterDF)
k6.avdist
```

```{r k7 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.7$cluster==1,],clusters.7$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.7$cluster==2,],clusters.7$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.7$cluster==3,],clusters.7$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.7$cluster==4,],clusters.7$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.7$cluster==5,],clusters.7$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.7$cluster==6,],clusters.7$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.7$cluster==7,],clusters.7$centers[7,,drop=FALSE])
# Get average distance
k7.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7)/nrow(clusterDF)
k7.avdist
```

```{r k8 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.8$cluster==1,],clusters.8$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.8$cluster==2,],clusters.8$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.8$cluster==3,],clusters.8$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.8$cluster==4,],clusters.8$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.8$cluster==5,],clusters.8$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.8$cluster==6,],clusters.8$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.8$cluster==7,],clusters.8$centers[7,,drop=FALSE])
dist.8 <- myDist(clusterDF[clusters.8$cluster==8,],clusters.8$centers[8,,drop=FALSE])
# Get average distance
k8.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7, dist.8)/nrow(clusterDF)
k8.avdist
```

```{r k9 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.9$cluster==1,],clusters.9$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.9$cluster==2,],clusters.9$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.9$cluster==3,],clusters.9$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.9$cluster==4,],clusters.9$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.9$cluster==5,],clusters.9$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.9$cluster==6,],clusters.9$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.9$cluster==7,],clusters.9$centers[7,,drop=FALSE])
dist.8 <- myDist(clusterDF[clusters.9$cluster==8,],clusters.9$centers[8,,drop=FALSE])
dist.9 <- myDist(clusterDF[clusters.9$cluster==9,],clusters.9$centers[9,,drop=FALSE])
# Get average distance
k9.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7, dist.8, dist.9)/nrow(clusterDF)
k9.avdist
```

```{r k10 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.10$cluster==1,],clusters.10$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.10$cluster==2,],clusters.10$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.10$cluster==3,],clusters.10$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.10$cluster==4,],clusters.10$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.10$cluster==5,],clusters.10$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.10$cluster==6,],clusters.10$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.10$cluster==7,],clusters.10$centers[7,,drop=FALSE])
dist.8 <- myDist(clusterDF[clusters.10$cluster==8,],clusters.10$centers[8,,drop=FALSE])
dist.9 <- myDist(clusterDF[clusters.10$cluster==9,],clusters.10$centers[9,,drop=FALSE])
dist.10 <- myDist(clusterDF[clusters.10$cluster==10,],clusters.10$centers[10,,drop=FALSE])
# Get average distance
k10.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7, dist.8, dist.9, dist.10)/nrow(clusterDF)
k10.avdist
```

```{r k11 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.11$cluster==1,],clusters.11$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.11$cluster==2,],clusters.11$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.11$cluster==3,],clusters.11$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.11$cluster==4,],clusters.11$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.11$cluster==5,],clusters.11$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.11$cluster==6,],clusters.11$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.11$cluster==7,],clusters.11$centers[7,,drop=FALSE])
dist.8 <- myDist(clusterDF[clusters.11$cluster==8,],clusters.11$centers[8,,drop=FALSE])
dist.9 <- myDist(clusterDF[clusters.11$cluster==9,],clusters.11$centers[9,,drop=FALSE])
dist.10 <- myDist(clusterDF[clusters.11$cluster==10,],clusters.11$centers[10,,drop=FALSE])
dist.11 <- myDist(clusterDF[clusters.11$cluster==11,],clusters.11$centers[11,,drop=FALSE])
# Get average distance
k11.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7, dist.8, dist.9, dist.10, dist.11)/nrow(clusterDF)
k11.avdist
```

```{r k12 dist, echo = TRUE}
# Get distances from each cluster
dist.1 <- myDist(clusterDF[clusters.12$cluster==1,],clusters.12$centers[1,,drop=FALSE])
dist.2 <- myDist(clusterDF[clusters.12$cluster==2,],clusters.12$centers[2,,drop=FALSE])
dist.3 <- myDist(clusterDF[clusters.12$cluster==3,],clusters.12$centers[3,,drop=FALSE])
dist.4 <- myDist(clusterDF[clusters.12$cluster==4,],clusters.12$centers[4,,drop=FALSE])
dist.5 <- myDist(clusterDF[clusters.12$cluster==5,],clusters.12$centers[5,,drop=FALSE])
dist.6 <- myDist(clusterDF[clusters.12$cluster==6,],clusters.12$centers[6,,drop=FALSE])
dist.7 <- myDist(clusterDF[clusters.12$cluster==7,],clusters.12$centers[7,,drop=FALSE])
dist.8 <- myDist(clusterDF[clusters.12$cluster==8,],clusters.12$centers[8,,drop=FALSE])
dist.9 <- myDist(clusterDF[clusters.12$cluster==9,],clusters.12$centers[9,,drop=FALSE])
dist.10 <- myDist(clusterDF[clusters.12$cluster==10,],clusters.12$centers[10,,drop=FALSE])
dist.11 <- myDist(clusterDF[clusters.12$cluster==11,],clusters.12$centers[11,,drop=FALSE])
dist.12 <- myDist(clusterDF[clusters.12$cluster==12,],clusters.12$centers[12,,drop=FALSE])
# Get average distance
k12.avdist <- sum(dist.1, dist.2, dist.3, dist.4, dist.5, dist.6, dist.7, dist.8, dist.9, dist.10, dist.11, dist.12)/nrow(clusterDF)
k12.avdist
```

#### Calculate this average distance from the center of each cluster for each value of k and plot it as a line chart where k is the x-axis and the average distance is the y-axis.

```{r plot dist, echo = TRUE}
k.values <- 2:12
avDist <- c(k2.avdist, k3.avdist, k4.avdist, k5.avdist, k6.avdist, k7.avdist, k8.avdist, k9.avdist, k10.avdist, k11.avdist, k12.avdist)
k.dist <- data.frame(k.values, avDist)

ggplot(k.dist, mapping = aes(k.values, avDist)) + geom_line() + geom_point() + labs(x = "k-Values", y = "Avg Distance", title = "Average distance to cluster center for each k value")
```

The results seem to vary slightly depending on the random subset of data used to determine clusters, but with the seed set to 42, I would say that the "elbow" point would appear to be around k = 7 because that is where the chart seems to stop dropping and start to really level out. There are many ways to determine an elbow point so it seems somewhat arbitrary (again, especially considering the built-in variation due to randomness).

As a side note, I started trying to figure out a cleaner way to calculate average distance from cluster center for each value of k by using a function with a built-in loop. I wasn't able to get it working initially and between life, work, and school I ran out of time to figure it out and implement it, but I'm sure it's possible.



